### 2016-05-04 p02-ch01-2014/paper.tex
Fix ref/label for Chapter 03

\section{Introduction}
\label{sec:introduction}
Java is beyond doubt one of the mainstream object oriented programming languages 
that supports a \emph{programming to interfaces} discipline \cite{oop:patterns,oop:survey}.
Through the years, Java has evolved from a mere programming language to a huge platform to drive and envision standards for mission-critical business applications.
Moreover, the Java language itself has evolved in these years to support its community with new language features and standards.
One of the noticeable domains of focus in the past decade has been distribution and concurrency in research and application.
This has led to valuable research results and numerous libraries and frameworks with an attempt to provide distribution and concurrency at the level of Java language.
However, it is widely recognized that the thread-based model of concurrency in Java that is a well-known approach is not appropriate for realizing distributed systems because of its inherent synchronous communication model.
On the other hand, the event-driven actor model of concurrency introduced by Hewitt \cite{Hewitt69} is 
a powerful concept for modeling distributed and concurrent
systems \cite{Agha90,agha97}. 
Different extensions of actors are proposed in several domains and are claimed to be the most suitable model of computation for many  applications \cite{Hewitt07-commitment}. 
Examples of these domains include designing embedded systems \cite{LeeACtorEmbedded03,LeeLN09}, 
wireless sensor networks \cite{CheongSensor05}, multi-core programming \cite{KarmaniSA09} 
and delivering cloud services through SaaS or PaaS \cite{Chang:DAIS07}.
This model of concurrent computation forms the basis of the
programming languages Erlang \cite{Armstrong10Erlang} and Scala
\cite{haller09tcs} that have recently gained in popularity, in part
due to their support for scalable concurrency.  Moreover, based on the
Java language itself, there are numerous libraries  that
provide  an implementation of an actor-based programming model.

@@

The paper continues as follows: 
in Section \ref{sec:relatedwork}, we briefly discuss a set of related works on actors and concurrent models especially on JVM platform. 
Section \ref{sec:example} presents an example that we use throughout the paper, we start to model the example using a library. 
Section \ref{sec:actor:programming} briefly introduces a concurrent modeling language and implements the example.
Section \ref{sec:j8:features} briefly discusses Java 8 features that this works uses for implementation.
Section \ref{sec:j8:actors:realization} presents how an actor model maps into programming in Java~8.
Section \ref{sec:impl:arch} discusses in detail the implementation architecture of the actor API.
Section \ref{sec:experiments} discusses how a number of benchmarks were performed for the implementation of the API and how they compare with current related works.
Section \ref{sec:conclusion} concludes the paper and discusses the future work.

\section{Related Work}
\label{sec:relatedwork}

@@

\section{State of the Art: An example}
\label{sec:example}
In the following, we illustrate the state of the art in actor programming by means of a simple example using
the Akka~\cite{akka} library which features asynchronous messaging and which is used to program actors in both Scala and Java.
We want to model in Akka an ``asynchronous ping-pong match'' between two actors represented by the 
two interfaces \jtt{IPing} and \jtt{IPong} which are depicted in Listings \ref{lst:ping:interface} and \ref{lst:pong:interface}.
An asynchronous call by the actor implementing the \jtt{IPong} interface of the \jtt{ping} method of the actor
implementing the \jtt{IPing} interface should generate an asynchronous call of the \jtt{pong} method of the callee,
and vice versa.
We intentionally design \jtt{ping} and \jtt{pong} methods to take  arguments in order to demonstrate 
how method arguments may affect the use of an actor model in an object-oriented style.

@@

To model an actor in Akka by a class, say  \jtt{Ping},  with interface \jtt{IPing}, this class is required  \emph{both} to \emph{extend}
a given pre-defined  class \jtt{UntypedActor}  and \emph{implement} the interface \jtt{IPing}, 
as depicted in Listings \ref{lst:ping:akka} and \ref{lst:pong:akka}.
The class \jtt{UntypedActor} provides two Akka framework methods \jtt{tell} and \jtt{onReceive} 
which are used to enqueue and dequeue asynchronous messages.
An asynchronous call to, for example,  the method \jtt{ping} then can be modeled by passing a user-defined  encoding of this call,
in this case by prefixing the string argument with the string ``pinged'', to a (synchronous) call
of the \jtt{tell} method which results in enqueuing the message.
In case this message is dequeued
the implementation of the \jtt{onReceive} method as provided by the \jtt{Ping} class then calls the \jtt{ping} method.

@@

Access to the sender of the message in Akka is provided by \jtt{sender()}.
In the main method as described in Listing \ref{lst:main:akka} we show how the initialize and start the ping/pong match.
Note that a reference to the ``pong'' actor is passed to the ``ping'' actor.

@@

\section{Actor Programming in Java}
\label{sec:actor:programming}

@@

Asynchronous method calls use futures as dynamically generated references  to return values.
The execution of a method can be (temporarily) suspended by release statements which give rise to a form
of  cooperative scheduling of method invocations inside concurrent (active) objects.
Release statements can be conditional (e.g., checking a future for the return value).
Listings \ref{lst:ping:abs}, \ref{lst:pong:abs} and \ref{lst:main:ABS} present an implementation of ping-pong example in ABS.
By means of the statement on line 6 of Listing \ref{lst:ping:abs}
a ``ping'' object  directly calls asynchronously the \jtt{pong} method  of its ``pong'' object, and vice versa.
Such a call is stored in the message queue and the called method is executed when the message is dequeued.
Note that variables in ABS are declared by interfaces.
In ABS, \jtt{Unit} is similar to \jtt{void} in Java.

@@

\section{Java 8 Features}
\label{sec:j8:features}

@@

\section{Modeling actors in Java~8}
\label{sec:j8:actors:realization}

@@

\paragraph{The \jtt{Actor} Interface}
We introduce an interface to model actors  using Java~8 features discussed in Section~\ref{sec:j8:features}.
Implementing an interface in Java means that the object exposes public APIs specified by the interface that is considered the behavior of the object.
Interface implementation is opposed to inheritance extension in which the object is 
possibly forced to expose behavior that may not be part of its intended interface.  
Using an interface for an actor  allows an object to preserve its own interfaces, and second, 
it allows for multiple interfaces to be implemented and composed.

@@

The \jtt{Actor} interface (Listings \ref{lst:j8:run} and \ref{lst:j8:enq})
provides a set of \bfjtt{default} methods, namely the \jtt{run} and \jtt{send} methods,  which the implementing classes do not need to re-implement.
This interface further  encapsulates a queue of messages that supports concurrent features of Java API
\footnote{Such API includes usage of different interfaces and classes in \jtt{java.util.concurrent} package \cite{jsr166}.
The concurrent Java API supports blocking and synchronization features in a high-level that is abstracted from the user.}.
We distinguish two types of messages:
messages that are  not expected to generate  any result and messages that are  expected to generate a result captured by a future value; i.e. 
an instance of \jtt{Future} in Java~ 8.
The first kind of messages are modeled as instances of \jtt{Runnable} and the second kind are modeled instances of \jtt{Callable}.
The default \jtt{run} method then takes a message from the queue, checks its type and executes the message correspondingly.
On the other hand, the default (overloaded) \jtt{send} method stores the sent message and creates a future
which is returned to the caller, in case of an instance of \jtt{Callable}.

@@

\paragraph{Modeling Asynchronous Messages} 
\label{par:asynchronous_messages}
We model an asynchronous call 
$$
\jtt{Future<V>} \; f = e_0\; !\; m(e_1,\ldots,e_n)
$$
to a method in ABS by the Java~8 code snippet of Listing \ref{lst:j8:call0}.
The final local variables $\jtt{u}_1$, $\ldots$, $\jtt{u}_n$ (of the caller) are used to store the values of
the Java~8 expressions $\jtt{e}_1$, $\ldots$, $\jtt{e}_n$ corresponding to the actual parameters  $e_1,\ldots,e_n$.
The types \jtt{T}$_i$, $i=1,\ldots,n$,  are the corresponding Java~8 types of  $e_i$, $i=1,\ldots,n$.
\begin{center}
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[mathescape,caption=Async messages with futures,label=lst:j8:call0]
final T$_1$ u$_1$ = e$_1$;
. . . 
final T$_n$ u$_n$ = e$_n$;
Future<V> v = e$_0$.send(
  () $\to$ { return m(u$_1$,$\ldots$,u$_n$); }
);
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[mathescape,caption=Async messages w/o futures,label=lst:j8:call1]
final T$_1$ u$_1$ = e$_1$;
. . . 
final T$_n$ u$_n$ = e$_n$;
e$_0$.send(
  { () $\to$ m(u$_1$,$\ldots$,u$_n$); }
);
\end{lstlisting}
\end{minipage}
\end{center}

The lambda expression  which encloses the above method invocation  is an instance of the  functional interface; e.g. \jtt{Callable}.
Note that the generated object which  represents the  lambda expression will contain the local context of the caller of the method ``$m$'' (including the local variables storing the values of the expressions $e_1,\ldots,e_n$), which will be restored upon execution of the lambda expression. 
Listing \ref{lst:j8:call1} models an asynchronous call to a method without a return value.

As an example, Listings \ref{lst:ping} and \ref{lst:pong} present the running ping/pong example, using the above API.
The main program to use ping and pong implementation is presented in Listing~\ref{lst:main:api}.

@@

% In Listing \ref{lst:ping}, the implementation uses a reference to another actor; namely \jtt{pong}.
% It sends a message to the actor, captures the result into a future value and waits for the result.
% The processing of the message in \jtt{pong} leads to the invocation of the method ``pong'' 
% using the provided argument and returns the result (Listing \ref{lst:pong}).
% 
As demonstrated in the above examples, the ``ping'' and ``pong'' objects \emph{preserve} 
their own \emph{interfaces}  contrary to the example depicted in Section \ref{sec:example} 
in which the objects \emph{extend} a specific ``universal actor abstraction'' to inherit methods and behaviors to become an actor.
Further,  messages are processed \emph{generically} by the \jtt{run} method described in Listing \ref{lst:j8:run}.
Although, in the first place, sending an asynchronous may look like to be able to change the recipient actor's state, this is not correct.
The variables that can be used in a lambda expression are \emph{effectively} final. 
In other words, in the context of a lambda expression, the recipient actor only provides a snapshot view of its state that cannot be changed.
This prevents abuse of lambda expressions to change the receiver's state.

@@

\paragraph{Modeling Cooperative Scheduling}
The ABS statement $\bfjtt{await}\; g$, where $g$ is a boolean guard, allows an active object to preempt the current method and schedule another one.
% We describe the  translation of await statements in the context of a general translation of statements in the ABS language.
We model cooperative scheduling by means of a call to the \jtt{await} method in Listing~\ref{lst:j8:await}.
Note that the preempted process is thus passed as an additional parameter and as such queued in case the guard is false,
otherwise it is executed.
Moreover, the generation of the continuation of the process is an optimization task for the code generation process to prevent code duplication.
\begin{center}
\begin{minipage}[t]{0.48\textwidth}
% \begin{lstlisting}[mathescape, caption=Java~8 conceptual await,label=lst:j8:await]
%  this.await($b$, $()\rightarrow \{S\}$)
%  // S is the continuation
% \end{lstlisting}
\begin{lstlisting}[mathescape,caption=main in ABS API,label=lst:main:api]
IPong pong = new Pong();
IPing ping = new Ping(pong);
ping.send(
  () -> ping.ping("")
);
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[mathescape,caption=Java~8 await implementation,label=lst:j8:await]
void await(final Boolean guard, 
           final Runnable cont) {
  if (!guard) { 
    this.send(() $\to$ 
      { this.await(guard, cont) })
  } else { cont.run() }
}
\end{lstlisting}
\end{minipage}
\end{center}

@@

% \textsl{Actor API in Java~8} 
% \label{par:actor_api_in_java_8}

\section{Implementation Architecture}
\label{sec:impl:arch}

Figure \ref{fig:actors:arch} presents the general layered architecture of the actor API in Java 8.
It consists of  three layers: the  routing layer  which
forms the foundation for the support of distribution and location transparency~\cite{KarmaniSA09} of actors,
the queuing layer  which allows for different implementations of the message queues, and finally, the processing layer
which implements the actual execution of the messages.
Each layer  allows for further customization by means of plugins.
The implementation is available at \surl{https://github.com/CrispOSS/abs-api}.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[scale=.6]{figs/abs-api-java8-2}  
\end{center}
\caption{Architecture of Actor API in Java~8}
\label{fig:actors:arch}
\end{figure}

@@

% We discuss the distribution of actors in this architecture in the next section.
% The implementation is available at \surl{https://github.com/CrispOSS/abs-api}.
% 
% \section{Distribution of Actors}
% \label{sec:distributtion}

We discuss next the distribution of actors in this architecture.
In the architecture presented  in Figure~\ref{fig:actors:arch}, each layer can be \emph{distributed} independently of another layer in a transparent way.
Not only the routing layer can provide distribution, the queue layer of the architecture may also be remote to take advantage of cluster storage for actor messages. 
A remote routing layer can provide access to actors transparently through standard naming or addresses.
% The implementation of remote actors is left as an the extension or plugin that is injected into the system.
% For instance, an extension may choose to provide remote actor access through HTTP layer; however, another may opt for a network layer implementation.
We exploit the main properties of actor model~\cite{agha97,Agha90} to distribute actors based on our implementation.
From a distributed perspective, the following are the main requirements for  distributing actors:
\begin{description}
\item[Reference Location Transparency] 
Actors communicate to one another using references. 
In an actor model, there is no in-memory object reference; however, every actor reference denotes a location by means of which the actor is accessible.
The reference location may be local to the calling actor or remote.
The reference location is \emph{physically} transparent for the calling actor.
\item[Communication Transparency]
A message $m$ from actor $A$ to actor $B$ may possibly lead to transferring $m$ over a network such that $B$ can process the message.
Thus, an actor model that supports distribution must provide a layer of remote communication among its actors that is transparent, i.e., 
when actor $A$ sends message $m$, the message is transparently transferred over the network to reach actor $B$.
For instance, actors existing in an HTTP container that transparently allows such communication.
Further, the API implementation is required to provide a mechanism for serialization of messages.
By default, every object in JVM cannot be assumed to be an instance of \jtt{java.io.Serializable}.
However, the API may enforce  that any remote actor should have the required actor classes in its JVM during runtime which
allows the use of the JVM's general object serialization
\footnote{Java Object Serialization Specification: \surl{http://docs.oracle.com/javase/8/docs/platform/serialization/spec/serialTOC.html}}
to send messages to remote actors and receive their responses.
Additionally, we model asynchronous messages with lambda expressions for which Java~8 supports serialization by specification
\footnote{Serialized Lambdas: \surl{http://docs.oracle.com/javase/8/docs/api/java/lang/invoke/SerializedLambda.html}}.
\item[Actor Provisioning]
During a life time of an actor, it may need to create new actors.
Creating actors in a local memory setting is straightforward. 
However, the local setting \emph{does} have a capacity of number of actors it can hold.
When an actor creates a new one, the new actor may actually be initialized in a remote resource.
When the resource is not available, it should be first provisioned.
However, this resource provisioning should be transparent to the actor and only the eventual result (the newly created actor) is visible.
\end{description}

We extend the ABS API to ABS Remote API\footnote{The implementation is available at \surl{https://github.com/CrispOSS/abs-api-remote}.} 
that provides the above properties for  actors in a seamless way.
A complete example of using the remote API has been developed\footnote{An example of ABS Remote API is available at \surl{https://github.com/CrispOSS/abs-api-remote-sample}.}. 
Expanding our ping-pong example in this paper, Listing~\ref{lst:ping:remote} and \ref{lst:pong:remote} present how a remote server of actors is created for the ping and pong actors.
In the following listings, \jtt{java.util.Properties} is used provide input parameters of the actor server;
namely, the address and the port that the actor server responds to.

@@

In Listing~\ref{lst:ping:remote}, a remote reference to a pong actor is created that exposes the \jtt{IPong} interface.
This interface is proxied
\footnote{Java Proxy: \surl{http://docs.oracle.com/javase/8/docs/api/java/lang/reflect/Proxy.html}} by the implementation 
to handle the remote communication with the actual pong actor in the other actor server.
This mechanism hides the communication details from the ping actor and as such allows the ping actor to use 
the same API to send a message to the pong actor (without even  knowing that the pong actor is actually remote).
When an actor is initialized in a distributed setting it transparently identifies its actor server and registers with it.
The above two listings are aligned with the similar main program presented in Listing~\ref{lst:main:api} that presents the same in a local setting.
The above two listings run in separate JVM instances and therefore do not share any objects.
In each JVM instance, it is required that both interfaces \jtt{IPing} and \jtt{IPong} are visible to the classpath; however, 
the ping actor server only needs to see \jtt{Ping} class in its classpath and similarly the pong actor server only needs 
to see \jtt{Pong} class in its classpath.

\section{Experiments}
\label{sec:experiments}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[scale=.45]{figs/benchmarks}  
\end{center}
\caption[Benchmarking comparison of ABS API and Akka]{
Benchmark results of comparing sampling time of message round trips in ABS API and Akka.
An example reading of above results is that the time shows for $p(90.0000)$ reads as ``message round trips were completed under $10{\mu}s$ for $90\%$ of the sent messages''.
The first two columns show the ``minimum'' and ``mean'' message round trip times in both implementations.
}
\label{fig:benchmarks}
\end{figure}

>>>>

\section{Introduction}
\label{ch03:sec:introduction}
Java is beyond doubt one of the mainstream object oriented programming languages 
that supports a \emph{programming to interfaces} discipline \cite{oop:patterns,oop:survey}.
Through the years, Java has evolved from a mere programming language to a huge platform to drive and envision standards for mission-critical business applications.
Moreover, the Java language itself has evolved in these years to support its community with new language features and standards.
One of the noticeable domains of focus in the past decade has been distribution and concurrency in research and application.
This has led to valuable research results and numerous libraries and frameworks with an attempt to provide distribution and concurrency at the level of Java language.
However, it is widely recognized that the thread-based model of concurrency in Java that is a well-known approach is not appropriate for realizing distributed systems because of its inherent synchronous communication model.
On the other hand, the event-driven actor model of concurrency introduced by Hewitt \cite{Hewitt69} is 
a powerful concept for modeling distributed and concurrent
systems \cite{Agha90,agha97}. 
Different extensions of actors are proposed in several domains and are claimed to be the most suitable model of computation for many  applications \cite{Hewitt07-commitment}. 
Examples of these domains include designing embedded systems \cite{LeeACtorEmbedded03,LeeLN09}, 
wireless sensor networks \cite{CheongSensor05}, multi-core programming \cite{KarmaniSA09} 
and delivering cloud services through SaaS or PaaS \cite{Chang:DAIS07}.
This model of concurrent computation forms the basis of the
programming languages Erlang \cite{Armstrong10Erlang} and Scala
\cite{haller09tcs} that have recently gained in popularity, in part
due to their support for scalable concurrency.  Moreover, based on the
Java language itself, there are numerous libraries  that
provide  an implementation of an actor-based programming model.

@@

The paper continues as follows: 
in Section \ref{ch03:sec:relatedwork}, we briefly discuss a set of related works on actors and concurrent models especially on JVM platform. 
Section \ref{ch03:sec:example} presents an example that we use throughout the paper, we start to model the example using a library. 
Section \ref{ch03:sec:actor:programming} briefly introduces a concurrent modeling language and implements the example.
Section \ref{ch03:sec:j8:features} briefly discusses Java 8 features that this works uses for implementation.
Section \ref{ch03:sec:j8:actors:realization} presents how an actor model maps into programming in Java~8.
Section \ref{ch03:sec:impl:arch} discusses in detail the implementation architecture of the actor API.
Section \ref{ch03:sec:experiments} discusses how a number of benchmarks were performed for the implementation of the API and how they compare with current related works.
Section \ref{ch03:sec:conclusion} concludes the paper and discusses the future work.

\section{Related Work}
\label{ch03:sec:relatedwork}

@@

\section{State of the Art: An example}
\label{ch03:sec:example}
In the following, we illustrate the state of the art in actor programming by means of a simple example using
the Akka~\cite{akka} library which features asynchronous messaging and which is used to program actors in both Scala and Java.
We want to model in Akka an ``asynchronous ping-pong match'' between two actors represented by the 
two interfaces \jtt{IPing} and \jtt{IPong} which are depicted in Listings \ref{ch03:lst:ping:interface} and \ref{ch03:lst:pong:interface}.
An asynchronous call by the actor implementing the \jtt{IPong} interface of the \jtt{ping} method of the actor
implementing the \jtt{IPing} interface should generate an asynchronous call of the \jtt{pong} method of the callee,
and vice versa.
We intentionally design \jtt{ping} and \jtt{pong} methods to take  arguments in order to demonstrate 
how method arguments may affect the use of an actor model in an object-oriented style.

@@

To model an actor in Akka by a class, say  \jtt{Ping},  with interface \jtt{IPing}, this class is required  \emph{both} to \emph{extend}
a given pre-defined  class \jtt{UntypedActor}  and \emph{implement} the interface \jtt{IPing}, 
as depicted in Listings \ref{ch03:lst:ping:akka} and \ref{ch03:lst:pong:akka}.
The class \jtt{UntypedActor} provides two Akka framework methods \jtt{tell} and \jtt{onReceive} 
which are used to enqueue and dequeue asynchronous messages.
An asynchronous call to, for example,  the method \jtt{ping} then can be modeled by passing a user-defined  encoding of this call,
in this case by prefixing the string argument with the string ``pinged'', to a (synchronous) call
of the \jtt{tell} method which results in enqueuing the message.
In case this message is dequeued
the implementation of the \jtt{onReceive} method as provided by the \jtt{Ping} class then calls the \jtt{ping} method.

@@

Access to the sender of the message in Akka is provided by \jtt{sender()}.
In the main method as described in Listing \ref{ch03:lst:main:akka} we show how the initialize and start the ping/pong match.
Note that a reference to the ``pong'' actor is passed to the ``ping'' actor.

@@

\section{Actor Programming in Java}
\label{ch03:sec:actor:programming}

@@

Asynchronous method calls use futures as dynamically generated references  to return values.
The execution of a method can be (temporarily) suspended by release statements which give rise to a form
of  cooperative scheduling of method invocations inside concurrent (active) objects.
Release statements can be conditional (e.g., checking a future for the return value).
Listings \ref{ch03:lst:ping:abs}, \ref{ch03:lst:pong:abs} and \ref{ch03:lst:main:ABS} present an implementation of ping-pong example in ABS.
By means of the statement on line 6 of Listing \ref{ch03:lst:ping:abs}
a ``ping'' object  directly calls asynchronously the \jtt{pong} method  of its ``pong'' object, and vice versa.
Such a call is stored in the message queue and the called method is executed when the message is dequeued.
Note that variables in ABS are declared by interfaces.
In ABS, \jtt{Unit} is similar to \jtt{void} in Java.

@@

\section{Java 8 Features}
\label{ch03:sec:j8:features}

@@

\section{Modeling actors in Java~8}
\label{ch03:sec:j8:actors:realization}

@@

\paragraph{The \jtt{Actor} Interface}
We introduce an interface to model actors  using Java~8 features discussed in Section~\ref{ch03:sec:j8:features}.
Implementing an interface in Java means that the object exposes public APIs specified by the interface that is considered the behavior of the object.
Interface implementation is opposed to inheritance extension in which the object is 
possibly forced to expose behavior that may not be part of its intended interface.  
Using an interface for an actor  allows an object to preserve its own interfaces, and second, 
it allows for multiple interfaces to be implemented and composed.

@@

The \jtt{Actor} interface (Listings \ref{ch03:lst:j8:run} and \ref{ch03:lst:j8:enq})
provides a set of \bfjtt{default} methods, namely the \jtt{run} and \jtt{send} methods,  which the implementing classes do not need to re-implement.
This interface further  encapsulates a queue of messages that supports concurrent features of Java API
\footnote{Such API includes usage of different interfaces and classes in \jtt{java.util.concurrent} package \cite{jsr166}.
The concurrent Java API supports blocking and synchronization features in a high-level that is abstracted from the user.}.
We distinguish two types of messages:
messages that are  not expected to generate  any result and messages that are  expected to generate a result captured by a future value; i.e. 
an instance of \jtt{Future} in Java~ 8.
The first kind of messages are modeled as instances of \jtt{Runnable} and the second kind are modeled instances of \jtt{Callable}.
The default \jtt{run} method then takes a message from the queue, checks its type and executes the message correspondingly.
On the other hand, the default (overloaded) \jtt{send} method stores the sent message and creates a future
which is returned to the caller, in case of an instance of \jtt{Callable}.

@@

\paragraph{Modeling Asynchronous Messages} 
\label{ch03:par:asynchronous_messages}
We model an asynchronous call 
$$
\jtt{Future<V>} \; f = e_0\; !\; m(e_1,\ldots,e_n)
$$
to a method in ABS by the Java~8 code snippet of Listing \ref{ch03:lst:j8:call0}.
The final local variables $\jtt{u}_1$, $\ldots$, $\jtt{u}_n$ (of the caller) are used to store the values of
the Java~8 expressions $\jtt{e}_1$, $\ldots$, $\jtt{e}_n$ corresponding to the actual parameters  $e_1,\ldots,e_n$.
The types \jtt{T}$_i$, $i=1,\ldots,n$,  are the corresponding Java~8 types of  $e_i$, $i=1,\ldots,n$.
\begin{center}
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[mathescape,caption=Async messages with futures,label=lst:j8:call0]
final T$_1$ u$_1$ = e$_1$;
. . . 
final T$_n$ u$_n$ = e$_n$;
Future<V> v = e$_0$.send(
  () $\to$ { return m(u$_1$,$\ldots$,u$_n$); }
);
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[mathescape,caption=Async messages w/o futures,label=lst:j8:call1]
final T$_1$ u$_1$ = e$_1$;
. . . 
final T$_n$ u$_n$ = e$_n$;
e$_0$.send(
  { () $\to$ m(u$_1$,$\ldots$,u$_n$); }
);
\end{lstlisting}
\end{minipage}
\end{center}

The lambda expression  which encloses the above method invocation  is an instance of the  functional interface; e.g. \jtt{Callable}.
Note that the generated object which  represents the  lambda expression will contain the local context of the caller of the method ``$m$'' (including the local variables storing the values of the expressions $e_1,\ldots,e_n$), which will be restored upon execution of the lambda expression. 
Listing \ref{ch03:lst:j8:call1} models an asynchronous call to a method without a return value.

As an example, Listings \ref{ch03:lst:ping} and \ref{ch03:lst:pong} present the running ping/pong example, using the above API.
The main program to use ping and pong implementation is presented in Listing~\ref{ch03:lst:main:api}.

@@

% In Listing \ref{ch03:lst:ping}, the implementation uses a reference to another actor; namely \jtt{pong}.
% It sends a message to the actor, captures the result into a future value and waits for the result.
% The processing of the message in \jtt{pong} leads to the invocation of the method ``pong'' 
% using the provided argument and returns the result (Listing \ref{ch03:lst:pong}).
% 
As demonstrated in the above examples, the ``ping'' and ``pong'' objects \emph{preserve} 
their own \emph{interfaces}  contrary to the example depicted in Section \ref{ch03:sec:example} 
in which the objects \emph{extend} a specific ``universal actor abstraction'' to inherit methods and behaviors to become an actor.
Further,  messages are processed \emph{generically} by the \jtt{run} method described in Listing \ref{ch03:lst:j8:run}.
Although, in the first place, sending an asynchronous may look like to be able to change the recipient actor's state, this is not correct.
The variables that can be used in a lambda expression are \emph{effectively} final. 
In other words, in the context of a lambda expression, the recipient actor only provides a snapshot view of its state that cannot be changed.
This prevents abuse of lambda expressions to change the receiver's state.

@@

\paragraph{Modeling Cooperative Scheduling}
The ABS statement $\bfjtt{await}\; g$, where $g$ is a boolean guard, allows an active object to preempt the current method and schedule another one.
% We describe the  translation of await statements in the context of a general translation of statements in the ABS language.
We model cooperative scheduling by means of a call to the \jtt{await} method in Listing~\ref{ch03:lst:j8:await}.
Note that the preempted process is thus passed as an additional parameter and as such queued in case the guard is false,
otherwise it is executed.
Moreover, the generation of the continuation of the process is an optimization task for the code generation process to prevent code duplication.
\begin{center}
\begin{minipage}[t]{0.48\textwidth}
% \begin{lstlisting}[mathescape, caption=Java~8 conceptual await,label=lst:j8:await]
%  this.await($b$, $()\rightarrow \{S\}$)
%  // S is the continuation
% \end{lstlisting}
\begin{lstlisting}[mathescape,caption=main in ABS API,label=lst:main:api]
IPong pong = new Pong();
IPing ping = new Ping(pong);
ping.send(
  () -> ping.ping("")
);
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[mathescape,caption=Java~8 await implementation,label=lst:j8:await]
void await(final Boolean guard, 
           final Runnable cont) {
  if (!guard) { 
    this.send(() $\to$ 
      { this.await(guard, cont) })
  } else { cont.run() }
}
\end{lstlisting}
\end{minipage}
\end{center}

@@

% \textsl{Actor API in Java~8} 
% \label{ch03:par:actor_api_in_java_8}

\section{Implementation Architecture}
\label{ch03:sec:impl:arch}

Figure \ref{ch03:fig:actors:arch} presents the general layered architecture of the actor API in Java 8.
It consists of  three layers: the  routing layer  which
forms the foundation for the support of distribution and location transparency~\cite{KarmaniSA09} of actors,
the queuing layer  which allows for different implementations of the message queues, and finally, the processing layer
which implements the actual execution of the messages.
Each layer  allows for further customization by means of plugins.
The implementation is available at \surl{https://github.com/CrispOSS/abs-api}.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[scale=.6]{figs/abs-api-java8-2}  
\end{center}
\caption{Architecture of Actor API in Java~8}
\label{ch03:fig:actors:arch}
\end{figure}

@@

% We discuss the distribution of actors in this architecture in the next section.
% The implementation is available at \surl{https://github.com/CrispOSS/abs-api}.
% 
% \section{Distribution of Actors}
% \label{ch03:sec:distributtion}

We discuss next the distribution of actors in this architecture.
In the architecture presented  in Figure~\ref{ch03:fig:actors:arch}, each layer can be \emph{distributed} independently of another layer in a transparent way.
Not only the routing layer can provide distribution, the queue layer of the architecture may also be remote to take advantage of cluster storage for actor messages. 
A remote routing layer can provide access to actors transparently through standard naming or addresses.
% The implementation of remote actors is left as an the extension or plugin that is injected into the system.
% For instance, an extension may choose to provide remote actor access through HTTP layer; however, another may opt for a network layer implementation.
We exploit the main properties of actor model~\cite{agha97,Agha90} to distribute actors based on our implementation.
From a distributed perspective, the following are the main requirements for  distributing actors:
\begin{description}
\item[Reference Location Transparency] 
Actors communicate to one another using references. 
In an actor model, there is no in-memory object reference; however, every actor reference denotes a location by means of which the actor is accessible.
The reference location may be local to the calling actor or remote.
The reference location is \emph{physically} transparent for the calling actor.
\item[Communication Transparency]
A message $m$ from actor $A$ to actor $B$ may possibly lead to transferring $m$ over a network such that $B$ can process the message.
Thus, an actor model that supports distribution must provide a layer of remote communication among its actors that is transparent, i.e., 
when actor $A$ sends message $m$, the message is transparently transferred over the network to reach actor $B$.
For instance, actors existing in an HTTP container that transparently allows such communication.
Further, the API implementation is required to provide a mechanism for serialization of messages.
By default, every object in JVM cannot be assumed to be an instance of \jtt{java.io.Serializable}.
However, the API may enforce  that any remote actor should have the required actor classes in its JVM during runtime which
allows the use of the JVM's general object serialization
\footnote{Java Object Serialization Specification: \surl{http://docs.oracle.com/javase/8/docs/platform/serialization/spec/serialTOC.html}}
to send messages to remote actors and receive their responses.
Additionally, we model asynchronous messages with lambda expressions for which Java~8 supports serialization by specification
\footnote{Serialized Lambdas: \surl{http://docs.oracle.com/javase/8/docs/api/java/lang/invoke/SerializedLambda.html}}.
\item[Actor Provisioning]
During a life time of an actor, it may need to create new actors.
Creating actors in a local memory setting is straightforward. 
However, the local setting \emph{does} have a capacity of number of actors it can hold.
When an actor creates a new one, the new actor may actually be initialized in a remote resource.
When the resource is not available, it should be first provisioned.
However, this resource provisioning should be transparent to the actor and only the eventual result (the newly created actor) is visible.
\end{description}

We extend the ABS API to ABS Remote API\footnote{The implementation is available at \surl{https://github.com/CrispOSS/abs-api-remote}.} 
that provides the above properties for  actors in a seamless way.
A complete example of using the remote API has been developed\footnote{An example of ABS Remote API is available at \surl{https://github.com/CrispOSS/abs-api-remote-sample}.}. 
Expanding our ping-pong example in this paper, Listing~\ref{ch03:lst:ping:remote} and \ref{ch03:lst:pong:remote} present how a remote server of actors is created for the ping and pong actors.
In the following listings, \jtt{java.util.Properties} is used provide input parameters of the actor server;
namely, the address and the port that the actor server responds to.

@@

In Listing~\ref{ch03:lst:ping:remote}, a remote reference to a pong actor is created that exposes the \jtt{IPong} interface.
This interface is proxied
\footnote{Java Proxy: \surl{http://docs.oracle.com/javase/8/docs/api/java/lang/reflect/Proxy.html}} by the implementation 
to handle the remote communication with the actual pong actor in the other actor server.
This mechanism hides the communication details from the ping actor and as such allows the ping actor to use 
the same API to send a message to the pong actor (without even  knowing that the pong actor is actually remote).
When an actor is initialized in a distributed setting it transparently identifies its actor server and registers with it.
The above two listings are aligned with the similar main program presented in Listing~\ref{ch03:lst:main:api} that presents the same in a local setting.
The above two listings run in separate JVM instances and therefore do not share any objects.
In each JVM instance, it is required that both interfaces \jtt{IPing} and \jtt{IPong} are visible to the classpath; however, 
the ping actor server only needs to see \jtt{Ping} class in its classpath and similarly the pong actor server only needs 
to see \jtt{Pong} class in its classpath.

\section{Experiments}
\label{ch03:sec:experiments}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[scale=.45]{figs/benchmarks}  
\end{center}
\caption[Benchmarking comparison of ABS API and Akka]{
Benchmark results of comparing sampling time of message round trips in ABS API and Akka.
An example reading of above results is that the time shows for $p(90.0000)$ reads as ``message round trips were completed under $10{\mu}s$ for $90\%$ of the sent messages''.
The first two columns show the ``minimum'' and ``mean'' message round trip times in both implementations.
}
\label{ch03:fig:benchmarks}
\end{figure}
